---
title: "plomics"
output: 
    github_document:
        toc: true
        toc_depth: 3
editor_options: 
  chunk_output_type: console
---

A collection of functions for placental DNA methylation analysis.

## Install

```{r message = F, eval = F}
remotes::install_github('wvictor14/plomics')
```

```{r, message = F, warning = F, include = F}
library(plyr)
library(tidyr)
library(dplyr)
library(ggplot2)
```

## Functions

### lmmatrix

Computes pairwise linear models between several variables.

```{r message = F}
library(minfiData)
library(plomics)

# load example data
data(RGsetEx)

# calculate pcs on the data
betas <- getBeta(RGsetEx)
pc_obj <- prcomp(t(na.omit(betas)), center = T, scale = T)

# get pc scores for each sample
rotated <- pc_obj$x

#rsquared
rsq <- lmmatrix(dep = rotated,
                ind = as.data.frame(pData(RGsetEx)[,c('Sample_Group', 'age', 'sex', 'status')]))

#pvalue
pva <- lmmatrix(dep = rotated,
                ind = as.data.frame(pData(RGsetEx)[,c('Sample_Group', 'age', 'sex', 'status')]),
                metric = 'Pvalue')
##### plot
# reshape first
rsq_plot <- rsq %>% as.data.frame() %>% 
  
  # add dep variables
  mutate(dep = rownames(rsq)) %>%
  
  # reshape
  gather(PC, rsquared, -dep)

pva_plot <- pva %>% as.data.frame() %>% 
  
  # add dep variables
  mutate(dep = rownames(rsq)) %>%
  
  # reshape
  gather(PC, pval, -dep) %>%
  
  # pvalue categories
  mutate(pval_cat = case_when(
    pval > 0.05  ~ '> 0.05',
    pval < 0.05 & pval > 0.01 ~ '< 0.05',
    pval < 0.01 & pval > 0.001 ~ '< 0.01',
    pval < 0.001 ~ '< 0.001'
  ))
  
ggplot(rsq_plot, aes(x = PC, y = dep, fill = rsquared)) +
  geom_tile() + theme_bw() +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_fill_gradientn(colours=c("white", "#ffffcc", "#41b6c4", "#2c7fb8", "#253494"), 
                       breaks = c(0,0.5,1), limits = c(0,1), 
                       guide = guide_colorbar(frame.colour = "black", ticks.colour = "black")) 
ggplot(pva_plot, aes(x = PC, y = dep, fill = pval_cat)) +
  geom_tile() + theme_bw() +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_fill_manual(values = c('> 0.05' = 'white', '< 0.05' = '#fee8c8', 
                               '< 0.01' = '#fdbb84', '< 0.001' = '#e34a33'))
```

### pairTest

To test if covariates are confounding each other, we need to pairwise tests of independence between
covariates. If at least one covariate is numeric, we can use linear regression. Otherwise if both
covariates are categorical, then a chi squared test must be used.

```{r, warning = F}
variables <- data.frame(
  row = c(1, 1, 2, 2, 3, 3),
  column = c(1, 1, 1, 2, 2, 2),
  Sex = c('m', 'm', 'm', 'f', 'f', 'f'),
  age = c(18, 19, 18, 27, 30, 16),
  ethnicity = c('AF', 'AF', 'AF', 'EU', 'EU', 'AS')
)

tests <- pairtest(variables)
# make categories
tests <- tests %>% 
  mutate(pval_cat = if_else(p.value < 0.001, '< 0.001',
                            if_else(p.value < 0.01, '< 0.01',
                                    if_else(p.value < 0.05, '< 0.05', '<1'))))
tests

# plot heatmap of associations
ggplot(tests, aes(x=Row, y = Column, fill = pval_cat)) +
  geom_tile(col = 'grey') + theme(panel.background = element_blank()) + 
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_fill_manual(values = c('> 0.05' = 'white', '< 0.05' = '#fee8c8', 
                               '< 0.01' = '#fdbb84', '< 0.001' = '#e34a33'))
```

### findsentrix

`findsentrix` takes a vector of sentrix IDs (chip identifiers) and searches a directory for IDAT 
files that match. The returned data frame contains two columns: (1) the sentrix ID (2) unique file 
paths for each idat that matches the sentrix ID. 

Here is an example using the robinson lab master sample sheet:

```{r}
# read in master sample sheet
ss <- readxl::read_xlsx('Z:/ROBLAB6 Infinium450k John/Master_Sample_Sheet.xlsx')

## specify idat directory
idat_dir <- 'Z:/ROBLAB6 Infinium450k John/EPIC Raw data/'

ss <- ss %>% 

# Take the first 6 EPIC samples
  dplyr::arrange(desc(Platform)) %>% 
  dplyr::slice(1:6) %>% 
  dplyr::select(Sample_Name, Sentrix_ID, Sentrix_Position) %>%
 
# create sentrix column
  dplyr::mutate(Sentrix = paste0(Sentrix_ID, '_', Sentrix_Position))
```

We created a sentrix ID by taking the chip serial number (confusingly named as "sentrix_ID") and 
pasting this to the position identifier ("Sentrix_Position"):

**Sentrix:** `r ss$Sentrix[1]`

**Sentrix Chip Number:** `r as.character(ss$Sentrix_ID[1])`

**Sentrix Position on chip:** `r ss$Sentrix_Position[1]`


Now we can use `findsentrix` to find the filepaths for idats that match each sentrix identifier:

```{r}
idatfiles <- findsentrix(sentrix = ss$Sentrix, directory = idat_dir)
idatfiles

# join all matches, retaining unmatched and multiple matched IDs
ss <- ss %>%
  dplyr::full_join(idatfiles, by = 'Sentrix')
```

Finally, we can load idats using this dataframe:

```{r}
## Now you can load in these samples with minfi::read.metharray.exp
rgset <- minfi::read.metharray.exp(targets = as.data.frame(ss), verbose = T)
rgset
```

